SULI Report!

### 2015-08-20 doc/suli-report-swe.tex
Revised the abstract, converted 0.09 deg^2 to 324 arcmin^2 throughout SULI report. #79, #81.

\begin{abstract}
As dark matter does not absorb or emit light, its distribution in the universe must be inferred through indirect effects such as the gravitational lensing of distant galaxies. While most sources are only weakly lensed, the systematic alignment of background galaxies around a foreground lens can constrain the mass of the lens which is largely in the form of dark matter. In this paper, I have implemented a framework to reconstruct all of the mass along lines of sight using a  best-case dark matter halo model in which the halo mass is known. This framework is then used to make predictions of the weak lensing of 3,240 generated source galaxies through a 0.09 deg$^2$ field of the Millennium Simulation. The lensed source ellipticities are characterized by the ellipticity-ellipticity and galaxy-mass correlation functions and compared to the same statistic for the intrinsic and ray-traced ellipticities. In the ellipticity-ellipticity correlation function, I find that the framework systematically under predicts the shear power on average by 117.0\% and fails to capture correlation from dark matter structure at scales larger than 1 arcminute. The model predicted galaxy-mass correlation function is in agreement with the ray-traced statistic from scales 0.2 to 0.7 arcminutes, but systematically underpredicts shear power at scales larger than 0.7 arcminutes on average by 23.2\%. Physical and computational shortcomings of the framework are discussed, as well as potential improvements for upcoming work.
\end{abstract}

@@

In this paper, I attempt to do the former by applying a simple dark matter halo model to reconstruct mass along lines of sight in the Millennium Simulation and predict the weakly lensed ellipticities of generated background sources. A brief introduction into the theory of galaxy ellipticities, the effects of strong and weak gravitational lensing, and dark matter halos is discussed in Section II. The implementation of a halo model on the Millennium Simulation and ellipticity predictions is described in Section III, and the results of the model on 3,240 galaxies on a field size of 0.09 deg$^2$ as well as a comparison of the predicted lensed ellipticities to the true ellipticities is given in Section IV.  Section V discusses limitations of the used \texttt{Pangloss} framework along with potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section VI.

@@

\subsection*{Generating Background Galaxies}
With a catalog of foreground galaxies chosen, a set of 3,240 background galaxies over a field of 0.09 deg$^2$ was generated, with density 10 per arcminute$^2$. As the The intrinsic orientation of each galaxy was sampled from a uniform distribution as, without lensing, there should be no preferred orientation due to the assumption of an isotropic universe. The magnitude of the galaxy ellipticities was sampled from a normal distribution with a standard deviation of 0.2, but any ellipticities with magnitude greater than one were re-sampled. Random ellipticity noise was sampled from a normal distribution with a standard deviation of 0.1 and added to the intrinsic ellipticities. Finally, each ellipticity was multiplied by $M=0.9$ to account for a multiplicative bias of 10\%.

@@

Using the presented methodology, 3,240 background galaxies in a 0.09 deg$^2$ subset of the (0,0,0,0) Millennium Simulation foreground catalog were generated and lensed by both the ray-traced shear and convergence maps as well as the \texttt{Pangloss} framework. With a lightcone radius of 2 arcminutes, the lightcones contained an average of ${950\pm70}$ foreground halos. Both sets of lensed ellipticities were characterized with correlation functions, as well as the intrinsic ellipticities of the sources before lensing.

@@

While the current \texttt{Pangloss} framework can handle the lensing predictions of 3,240 in a 0.09 deg$^2$ field comfortably on a single processor, the goal is to scale up the framework to make predictions for background sources of the same number density in a 100 deg$^2$ field; this will require 3.6 million lightcones, each containing up to 1,000 foreground halos. Luckily, the prediction is trivially parallelizable as the shear and convergence calculation of each lightcone is independent of all other lightcones. This makes GPUs an attractive candidate for future work, as it would only take 360 GPUs with 10,000 threads each to carry out the prediction. Additionally, the (??) MB of RAM per lightcone is sufficiently small to fit 10,000 lightcones on a GPU.

@@

In this paper, a simple halo model was used to reconstruct all the mass along lines of sight in the Millennium Simulation to make predictions of how foreground mass weakly lensed the ellipticities of 3,240 generated background sources accross a field of 0.09 deg$^2$. The lensed ellipticities were characterized globally using the ellipticity-ellipticity and galaxy-mass correlation functions and then compared with the same statistics from ray-traced data. There was a systematic-underprediction of shear power on all scales in the ellipticity-ellipticity correlation function as well as an absence of any predicted correlation on scales larger than 1 arcminute, suggesting missing features in the \texttt{Pangloss} framework. The galaxy-mass correlation function predicted by the halo model was in agreement with the ray-traced correlation function on scales from 0.2 to 0.7 arcminutes, but significantly overpredicted the shear power on smaller scales due to strong lensing effects. Now that a proof of concept has been demonstrated, there is work being done to scale up the \texttt{Pangloss} framework to make lensing predictions over 100 deg$^2$ using far more computational resources. While the results are encouraging for such a simple model, many strong and unrealistic assumptions were made in the methodology that limit its use in observational surveys. In the future, a similar analysis should be made after relaxing many of the assumptions to see how different the predictions by the halo model are from this best-case scenario.

>>>>

\begin{abstract}
As dark matter does not absorb or emit light, its distribution in the universe must be inferred through indirect effects such as the gravitational lensing of distant galaxies. While most sources are only weakly lensed, the systematic alignment of background galaxies around a foreground lens can constrain the mass of the lens which is largely in the form of dark matter. In this paper, I have implemented a framework to reconstruct all of the mass along lines of sight using a  best-case dark matter halo model in which the halo mass is known. This framework is then used to make predictions of the weak lensing of 3,240 generated source galaxies through a 324 arcmin$^2$ field of the Millennium Simulation. The lensed source ellipticities are characterized by the ellipticity-ellipticity and galaxy-mass correlation functions and compared to the same statistic for the intrinsic and ray-traced ellipticities. In the ellipticity-ellipticity correlation function, I find that the framework systematically under predicts the shear power on average by a factor of 2.2 and fails to capture correlation from dark matter structure at scales larger than 1 arcminute. The model predicted galaxy-mass correlation function is in agreement with the ray-traced statistic from scales 0.2 to 0.7 arcminutes, but systematically underpredicts shear power at scales larger than 0.7 arcminutes on average by a factor of 1.2. Physical and computational shortcomings of the framework are discussed, as well as potential improvements for upcoming work.
\end{abstract}

@@

In this paper, I attempt to do the former by applying a simple dark matter halo model to reconstruct mass along lines of sight in the Millennium Simulation and predict the weakly lensed ellipticities of generated background sources. A brief introduction into the theory of galaxy ellipticities, the effects of strong and weak gravitational lensing, and dark matter halos is discussed in Section II. The implementation of a halo model on the Millennium Simulation and ellipticity predictions is described in Section III, and the results of the model on 3,240 galaxies on a field size of 324 arcmin$^2$ as well as a comparison of the predicted lensed ellipticities to the true ellipticities is given in Section IV.  Section V discusses limitations of the used \texttt{Pangloss} framework along with potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section VI.

@@

\subsection*{Generating Background Galaxies}
With a catalog of foreground galaxies chosen, a set of 3,240 background galaxies over a field of 324 arcmin$^2$ was generated, with density 10 per arcminute$^2$. As the The intrinsic orientation of each galaxy was sampled from a uniform distribution as, without lensing, there should be no preferred orientation due to the assumption of an isotropic universe. The magnitude of the galaxy ellipticities was sampled from a normal distribution with a standard deviation of 0.2, but any ellipticities with magnitude greater than one were re-sampled. Random ellipticity noise was sampled from a normal distribution with a standard deviation of 0.1 and added to the intrinsic ellipticities. Finally, each ellipticity was multiplied by $M=0.9$ to account for a multiplicative bias of 10\%.

@@

Using the presented methodology, 3,240 background galaxies in a 324 arcmin$^2$ subset of the (0,0,0,0) Millennium Simulation foreground catalog were generated and lensed by both the ray-traced shear and convergence maps as well as the \texttt{Pangloss} framework. With a lightcone radius of 2 arcminutes, the lightcones contained an average of ${950\pm70}$ foreground halos. Both sets of lensed ellipticities were characterized with correlation functions, as well as the intrinsic ellipticities of the sources before lensing.

@@

While the current \texttt{Pangloss} framework can handle the lensing predictions of 3,240 in a 324 arcmin$^2$ field comfortably on a single processor, the goal is to scale up the framework to make predictions for background sources of the same number density in a 100 deg$^2$ field; this will require 3.6 million lightcones, each containing up to 1,000 foreground halos. Luckily, the prediction is trivially parallelizable as the shear and convergence calculation of each lightcone is independent of all other lightcones. This makes GPUs an attractive candidate for future work, as it would only take 360 GPUs with 10,000 threads each to carry out the prediction. Additionally, the (??) MB of RAM per lightcone is sufficiently small to fit 10,000 lightcones on a GPU.

@@

In this paper, a simple halo model was used to reconstruct all the mass along lines of sight in the Millennium Simulation to make predictions of how foreground mass weakly lensed the ellipticities of 3,240 generated background sources accross a field of 324 arcmin$^2$. The lensed ellipticities were characterized globally using the ellipticity-ellipticity and galaxy-mass correlation functions and then compared with the same statistics from ray-traced data. There was a systematic-underprediction of shear power on all scales in the ellipticity-ellipticity correlation function as well as an absence of any predicted correlation on scales larger than 1 arcminute, suggesting missing features in the \texttt{Pangloss} framework. The galaxy-mass correlation function predicted by the halo model was in agreement with the ray-traced correlation function on scales from 0.2 to 0.7 arcminutes, but significantly overpredicted the shear power on smaller scales due to strong lensing effects. Now that a proof of concept has been demonstrated, there is work being done to scale up the \texttt{Pangloss} framework to make lensing predictions over 100 deg$^2$ using far more computational resources. While the results are encouraging for such a simple model, many strong and unrealistic assumptions were made in the methodology that limit its use in observational surveys. In the future, a similar analysis should be made after relaxing many of the assumptions to see how different the predictions by the halo model are from this best-case scenario.
